### 2023-09-26 doc/_bayesian_inference.qmd
small fixes to section

As defined in @sec-fitness_model, our ultimate objective is to infer the vector
of relative fitness values
$$
\underline{s}^M = (s^{(1)}, s^{(2)}, \ldots, s^{(M)})^\dagger,
$${#eq-fitness_mut_vec}
where $^\dagger$ indicates the transpose. Our data consists of an $T \times B$
matrix $\underline{\underline{R}}$, where $B = M + N$ is the number of unique
barcodes given by the sum of the number of unique, relevant barcodes we care
about (hereafter referred to as mutant barcodes), $M$, and the number of unique
neutral barcodes, $N$, and $T$ is the number of time points where measurements
were taken. The data matrix is then of the form
$$
\underline{\underline{R}} = \begin{bmatrix}
- & \underline{r}_1 & - \\
- & \underline{r}_2 & - \\
 & \vdots & \\
- & \underline{r}_T & - \\
\end{bmatrix},
$${#eq-R_mat}
where each row $\underline{r}_t$ is a $B$-dimensional array containing the
raw barcode counts at time $t$. We can further split each vector
$\underline{r}_t$ into two vectors of the form
$$
\underline{r}_t = \begin{bmatrix}
\underline{r}_t^{N} \;
\underline{r}_t^{M}
\end{bmatrix},
$${#eq-r-vec_split}
i.e., the vector containing the neutral lineage barcode counts
$\underline{r}_t^{N}$ and the corresponding vector containing the mutant barcode
counts $\underline{r}_t^{M}$. Following the same logic, matrix
$\underline{\underline{R}}$ can be split into two matrices as
$$
\underline{\underline{R}} = \left[ 
\underline{\underline{R}}^N \; \underline{\underline{R}}^M
\right],
$${#eq-R-mat_split}
where $\underline{\underline{R}}^N$ is a $T \times N$ matrix with the
barcode reads time series for each neutral lineage and
$\underline{\underline{R}}^M$ is the equivalent $T \times M$ matrix for the
mutant lineages.

Our objective is to compute the joint probability distribution for all relative
fitness values given our data. We can express this joint posterior distribution
using Bayes theorem as
$$
\pi(\underline{s}^M \mid \underline{\underline{R}}) = \frac{
\pi(\underline{\underline{R}} \mid \underline{s}^M) 
\pi(\underline{s}^M)}
{\pi(\underline{\underline{R}})},
$${#eq-bayes_obj}
where hereafter $\pi(\cdot)$ defines a probability density function, unless
otherwise stated. When defining our statistical model, we need not to focus on
the denominator on the right-hand side of @eq-bayes_obj, and write
$$
\pi(\underline{s}^M \mid \underline{\underline{R}}) \propto
\pi(\underline{\underline{R}} \mid \underline{s}^M) 
\pi(\underline{s}^M)
$${#eq-bayes_obj_propto}
However, when implementing the model computationally, the normalization constant
on the right-hand side of @eq-bayes_obj must be computed via an integral of the
form
$$
\pi(\underline{\underline{R}}) = \int d\underline{s}^M
\pi(\underline{\underline{R}} \mid \underline{s}^M) 
\pi(\underline{s}^M),
$${#eq-bayes_evidence}
also known as a marginalization integral.

Although @eq-bayes_obj and @eq-bayes_obj_propto seem simple enough, recall that
@eq-fitness relates barcode frequency values and the population mean fitness to
the mutant relative fitness. Therefore, we must include these nuisance
parameters as part of our inference problem. To include these nuisance
parameters, let 
$$
\underline{\bar{s}}_T = (\bar{s}_1, \bar{s}_2, \ldots, \bar{s}_{T-1})^\dagger,
$${#eq-pop_fitness_vec}
be the vector containing the $T-1$ population mean fitness we compute from the
$T$ time points where measurements were taken. We have $T-1$ since the value
of any $\bar{s}_t$ requires time points $t$ and $t+1$. Furthermore, let the
matrix $\underline{\underline{F}}$ be a $T \times B$ matrix containing all
frequency values. As with @eq-R_mat, we can split $\underline{\underline{F}}$
into two matrices of the form
$$
\underline{\underline{F}} = \left[ 
\underline{\underline{F}}^N \; \underline{\underline{F}}^M
\right],
$${#eq-F-mat_split}
to separate the corresponding mutant and neutral barcode frequencies. With these
nuisance variables in hand, the full inference problem we must solve takes the
form
$$
\pi(
    \underline{s}^M, \underline{\bar{s}}_T, \underline{\underline{F}} \mid
    \underline{\underline{R}}
) \propto
\pi(
    \underline{\underline{R}} \mid
    \underline{s}^M, \underline{\bar{s}}_T, \underline{\underline{F}}
)
\pi(
    \underline{s}^M, \underline{\bar{s}}_T, \underline{\underline{F}}
).
$${#eq-bayes_full}
We can numerically integrate out all nuisance parameters to recover the marginal
distribution we care about, i.e.,
$$
\pi(\underline{s}^M \mid \underline{\underline{R}}) =
\int d^{T-1}\underline{\bar{s}}_T
\int d^{B}\underline{f}_1 \cdots
\int d^{B}\underline{f}_T
\;
\pi(
    \underline{s}^M, \underline{\bar{s}}_T, \underline{\underline{F}} \mid
    \underline{\underline{R}}
).
$${#eq-bayes_full_marginal}

@@

The left-hand side of @eq-bayes_full is extremely difficult to work with.
However, we can take advantage of the structure of our inference problem to
rewrite it in a more manageable form. Specifically, the statistical dependencies
of our observations and latent variables allow us to factorize the joint
distribution into the product of multiple conditional distributions. To gain
some intuition about this factorization, let us focus on the inference of the
population mean fitness values $\underline{\bar{s}}_T$. @eq-logfreq_neutral
relates the value of the population mean fitness to the neutral lineage
frequencies and nothing else. This suggests that when writing the posterior for
these population mean fitness parameters, we should be able to condition it only
on the neutral lineage frequency values, i.e., $\pi(\underline{\bar{s}}_T \mid
\underline{\underline{F}}^N)$. We point the reader to `[SI section]` for the
full mathematical details on this factorization. For our purpose here, it
suffices to say we can rewrite the joint probability distribution as a product
of conditional distributions of the form
$$
\pi(
    \underline{s}^M, \underline{\bar{s}}_T, \underline{\underline{F}} \mid
    \underline{\underline{R}}
) =
\pi(
    \underline{s}^M \mid \underline{\bar{s}}_T, \underline{\underline{F}}^M
)
\pi(
    \underline{\bar{s}}_T \mid \underline{\underline{F}}^N
)
\pi(\underline{\underline{F}} \mid \underline{\underline{R}}).
$${#eq-split_posterior}
Written in this form, @eq-split_posterior captures the three sources of 
uncertainty listed in @sec-fitness_model in each term. Starting from
right to left, the first term on the right-hand side of @eq-split_posterior
accounts for the uncertainty when inferring the frequency values given the
barcode reads. The second term accounts for the uncertainty in the values of
the mean population fitness at different time points. The last term accounts for
the uncertainty in the parameter we care about---the mutants' relative
fitnesses. We invite the reader to check the supplementary material, where we
work through these terms to build the posterior distribution to sample
from. We refer the reader to `[ref SI section]` for an extended description of
the model with specific functional forms for each term on the left-hand side of
@eq-split_posterior as well as `[ref SI on hierarchical]` where the model is
extended to account for multiple experimental replicates.

>>>>

As defined in @sec-fitness_model, our ultimate objective is to infer the vector
of relative fitness values
$$
\underline{s}^M = (s^{(1)}, s^{(2)}, \ldots, s^{(M)})^\dagger,
$${#eq-fitness_mut_vec}
where $^\dagger$ indicates the transpose. Our data consists of an $T \times B$
matrix $\underline{\underline{R}}$, where $B = M + N$ is the number of unique
barcodes given by the sum of the number of unique, relevant barcodes we care
about, $M$, and the number of unique neutral barcodes, $N$, and $T$ is the
number of growth cycles where measurements were taken. The data matrix is then
of the form
$$
\underline{\underline{R}} = \begin{bmatrix}
- & \underline{r}_1 & - \\
- & \underline{r}_2 & - \\
 & \vdots & \\
- & \underline{r}_T & - \\
\end{bmatrix},
$${#eq-R_mat}
where each row $\underline{r}_t$ is a $B$-dimensional array containing the raw
barcode counts at cycle $t$. We can further split each vector $\underline{r}_t$
into two vectors of the form
$$
\underline{r}_t = \begin{bmatrix}
\underline{r}_t^{N} \\
\underline{r}_t^{M}
\end{bmatrix},
$${#eq-r-vec_split}
i.e., the vector containing the neutral lineage barcode counts
$\underline{r}_t^{N}$ and the corresponding vector containing the mutant barcode
counts $\underline{r}_t^{M}$. Following the same logic, matrix
$\underline{\underline{R}}$ can be split into two matrices as
$$
\underline{\underline{R}} = \left[ 
\underline{\underline{R}}^N \; \underline{\underline{R}}^M
\right],
$${#eq-R-mat_split}
where $\underline{\underline{R}}^N$ is a $T \times N$ matrix with the barcode
reads time series for each neutral lineage and $\underline{\underline{R}}^M$ is
the equivalent $T \times M$ matrix for the non-neutral lineages.

Our objective is to compute the joint probability distribution for all relative
fitness values given our data. We can express this joint posterior distribution
using Bayes theorem as
$$
\pi(\underline{s}^M \mid \underline{\underline{R}}) = \frac{
\pi(\underline{\underline{R}} \mid \underline{s}^M) 
\pi(\underline{s}^M)}
{\pi(\underline{\underline{R}})},
$${#eq-bayes_obj}
where hereafter $\pi(\cdot)$ defines a probability density function, unless
otherwise stated. When defining our statistical model, we need not to focus on
the denominator on the right-hand side of @eq-bayes_obj. Thus, we can write
$$
\pi(\underline{s}^M \mid \underline{\underline{R}}) \propto
\pi(\underline{\underline{R}} \mid \underline{s}^M) 
\pi(\underline{s}^M).
$${#eq-bayes_obj_propto}
However, when implementing the model computationally, the normalization constant
on the right-hand side of @eq-bayes_obj must be computed. This can be done from
the definition of the model via an integral of the form
$$
\pi(\underline{\underline{R}}) = \int d^M \underline{s}^M
\pi(\underline{\underline{R}} \mid \underline{s}^M) 
\pi(\underline{s}^M),
$${#eq-bayes_evidence}
also known as a marginalization integral. Hereafter, differentials of the form
$d^n$ imply a $n$-dimensional integral.

Although @eq-bayes_obj and @eq-bayes_obj_propto seem simple enough, recall that
@eq-fitness relates barcode frequency values and the population mean fitness to
the mutant relative fitness. Therefore, we must include these nuisance
parameters as part of our inference problem. To include these nuisance
parameters, let 
$$
\underline{\bar{s}}_T = (\bar{s}_1, \bar{s}_2, \ldots, \bar{s}_{T-1})^\dagger,
$${#eq-pop_fitness_vec}
be the vector containing the $T-1$ population mean fitness we compute from the
$T$ time points where measurements were taken. We have $T-1$ since the value
of any $\bar{s}_t$ requires cycle numbers $t$ and $t+1$. Furthermore, let the
matrix $\underline{\underline{F}}$ be a $T \times B$ matrix containing all
frequency values. As with @eq-R-mat_split, we can split
$\underline{\underline{F}}$ into two matrices of the form
$$
\underline{\underline{F}} = \left[ 
\underline{\underline{F}}^N \; \underline{\underline{F}}^M
\right],
$${#eq-F-mat_split}
to separate the corresponding neutral and non-neutral barcode frequencies. With
these nuisance variables in hand, the full inference problem we must solve takes
the form
$$
\pi(
    \underline{s}^M, \underline{\bar{s}}_T, \underline{\underline{F}} \mid
    \underline{\underline{R}}
) \propto
\pi(
    \underline{\underline{R}} \mid
    \underline{s}^M, \underline{\bar{s}}_T, \underline{\underline{F}}
)
\pi(
    \underline{s}^M, \underline{\bar{s}}_T, \underline{\underline{F}}
).
$${#eq-bayes_full}
To recover the marginal distribution over the non-neutral barcodes relative
fitness values, we can numerically integrate out all nuisance parameters, i.e.,
$$
\pi(\underline{s}^M \mid \underline{\underline{R}}) =
\int d^{T-1}\underline{\bar{s}}_T
\int d^{B}\underline{f}_1 \cdots
\int d^{B}\underline{f}_T
\;
\pi(
    \underline{s}^M, \underline{\bar{s}}_T, \underline{\underline{F}} \mid
    \underline{\underline{R}}
).
$${#eq-bayes_full_marginal}

@@

The left-hand side of @eq-bayes_full is extremely difficult to work with.
However, we can take advantage of the structure of our inference problem to
rewrite it in a more manageable form. Specifically, the statistical dependencies
of our observations and latent variables allow us to factorize the joint
distribution into the product of multiple conditional distributions. To gain
some intuition about this factorization, let us focus on the inference of the
population mean fitness values $\underline{\bar{s}}_T$. @eq-logfreq_neutral
relates the value of the population mean fitness to the neutral lineage
frequencies and nothing else. This suggests that when writing the posterior for
these population mean fitness parameters, we should be able to condition it only
on the neutral lineage frequency values, i.e., $\pi(\underline{\bar{s}}_T \mid
\underline{\underline{F}}^N)$. We point the reader to @sec-bayes_def for the
full mathematical details on this factorization. For our purpose here, it
suffices to say we can rewrite the joint probability distribution as a product
of conditional distributions of the form
$$
\pi(
    \underline{s}^M, \underline{\bar{s}}_T, \underline{\underline{F}} \mid
    \underline{\underline{R}}
) =
\pi(
    \underline{s}^M \mid \underline{\bar{s}}_T, \underline{\underline{F}}^M
)
\pi(
    \underline{\bar{s}}_T \mid \underline{\underline{F}}^N
)
\pi(\underline{\underline{F}} \mid \underline{\underline{R}}).
$${#eq-split_posterior}
Written in this form, @eq-split_posterior captures the three sources of
uncertainty listed in @sec-fitness_model in each term. Starting from right to
left, the first term on the right-hand side of @eq-split_posterior accounts for
the uncertainty when inferring the frequency values given the barcode reads. The
second term accounts for the uncertainty in the values of the mean population
fitness at different time points. The last term accounts for the uncertainty in
the parameter we care about---the mutants' relative fitnesses. We refer the
reader to @sec-bayes_def for an extended description of the model with specific
functional forms for each term on the left-hand side of @eq-split_posterior as
well as the extension of the model to account for multiple experimental
replicates or hierarchical genotypes.
